---
title: "Reproducible Research Project 1"
output:
  html_document:
    df_print: paged
    theme: spacelab
    highlight: pygments
    number_sections: yes
  html_notebook: default
  pdf_document: 
    highlight: pygments
    number_sections: yes
---

This is Coursera Reproducible Research Project Week 2 Peer Review Assignement.


```{r, message = FALSE}

library(pander)

library(tidyverse)

library(xtable)

library(knitr)

library(gridExtra)

options(xtable.comment = FALSE)
```

```{r}
inline_hook <- function (x) {
  if (is.numeric(x)) {
    ## ifelse does a vectorized comparison
    ## If integer, print without decimal; otherwise print two places
    res <- ifelse(x == round(x),
      sprintf("%d", x),
      sprintf("%.2f", x)
    )
    paste(res, collapse = ", ")
  }
}

knit_hooks$set(inline = inline_hook)

```

# Loading and Preprocessing Data 

1. load the data
2. Process/transform the data

```{r}
activity <- read.csv("activity.csv")

pander(head (activity, 6))

table(is.na(activity$steps))

table(is.na(activity$date))

table(is.na(activity$interval))

```


3. remove the missing value 

```{r}

activity_NARemoved <- na.omit(activity)

```


# Mean Total Number of Steps Taken per Day

1. calculate the total number of steps per day

```{r}
DailyStep <- activity_NARemoved %>% group_by(date) %>% 
    summarise(Daily_Step = sum(steps))

sum(is.na(DailyStep))

pander(head(DailyStep,6))

DailyStep$Daily_Step <- as.numeric(DailyStep$Daily_Step)

```

2. Histogram: the total number of steps taken each day  
```{r}
DailyStepHist <- ggplot (DailyStep, aes(Daily_Step)) + 
    geom_histogram(fill ="blue",binwidth = 1000)+
    labs(title = "Histogram of the Total Number of Steps Taken Each Day", x = "Daily_Step Missing Values Removed")

DailyStepHist
```

3. Mean and Median of the total number of steps taken per day

The mean of the total number of steps taken per day with missing value is ``r as.numeric(mean(DailyStep$Daily_Step))``, and 

the median of the total number of streps taken per day with missing value is ``r as.numeric(median(DailyStep$Daily_Step))``.

```{r}
mean(DailyStep$Daily_Step)

median(DailyStep$Daily_Step)
```

# The Average Daily Activity Pattern

1. A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
five_min_steps <- activity_NARemoved %>% group_by(interval) %>% 
    summarise(StepsPerFiveMin = mean(steps))

five_min_steps_TimeSeries <- ggplot(five_min_steps,aes(interval,StepsPerFiveMin))+
    geom_line()+ylab("Number of Steps")+xlab("5-Minutes Interval")+
    ggtitle("Average Number of Steps Taken Every 5 Minutes")

five_min_steps_TimeSeries
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The maximum number of steps 

```{r}
Max_Steps <- five_min_steps[five_min_steps$StepsPerFiveMin == max(five_min_steps$StepsPerFiveMin),]

pander(Max_Steps)

```

# Imputing missing values

1. the total number of missing values in the dataset is ``r sum(is.na(activity$steps))``

```{r total number of missing values}
sum(is.na(activity$steps))

```

2. Assign mean to the missing values 

Calculate mean within `activity_NARemoved` data frame.

```{r calcualte mean value without missing value}
AverageByInterval <- activity_NARemoved %>% group_by(interval) %>%
    summarise(AverageSteps = mean(steps))

pander(head(AverageByInterval,6))

```

Prepare a data frame for missing value only

```{r Missing Value}
activityMissingValue <- activity %>% filter(is.na(steps)) %>% group_by(interval,date)

pander(head(activityMissingValue))

```

Assign mean values to the missing value using `merge function`
```{r}
AssignMissingValue <- merge(activityMissingValue, AverageByInterval, by="interval")

pander(head(AssignMissingValue,3))



```

```{r}
AssignMissingValue <- AssignMissingValue [, c(1,3,4)]

pander(head(AssignMissingValue,3))


```

All missing values have been imputed using the mean value. 

```{r rename data frame}
## rename data frame

colnames(AssignMissingValue) = c("interval","date","steps")

pander(head(AssignMissingValue,3))

```

Combine two dataframes

```{r Combine two dataset}
activity2 <- rbind(activity_NARemoved,AssignMissingValue)

pander(head(activity2,6))
```

check total observation in the activity2 - missing value filled adn activity- original data set

```{r}
dim(activity)

dim(activity2)
```

checking on number of NA

```{r}
sum(is.na(activity2$steps))
```

calculate the total steps per day

```{r}
activityFull <- activity2 %>% group_by(date) %>% summarise(Daily_Step = sum(steps))

pander(head(activityFull,6))
```
Mean and Median in the new dataset activity2 - no missing value. The mean after
missing value filled is ``r mean(activityFull$Daily_Step)``, and the median afer missing values filled is ``r median (activityFull$Daily_Step)``

```{r}

mean(activityFull$Daily_Step)

median (activityFull$Daily_Step)

```

Compare dataset with missing value filled to the dataset with missing value

```{r}

Item <- c("Mean","Median")
OriginalData <- c(mean(DailyStep$Daily_Step),median(DailyStep$Daily_Step))
ImputedData <- c(mean(activityFull$Daily_Step),median(activityFull$Daily_Step))
Compare <- data.frame(Item,OriginalData,ImputedData)

pander(Compare)

```

make a Histogram on the new dataset

```{r}
DailyStepHistFull <- ggplot (activityFull, aes(x=Daily_Step)) + 
    geom_histogram(fill ="blue",binwidth = 1000)+
    labs(title = "Histogram of the Total Number of Steps Taken Each Day", x = "Daily_Step Missing Value Imputed")

DailyStepHistFull
```

```{r}
grid.arrange(DailyStepHist,DailyStepHistFull,ncol=2)
```

# Differences in activity patterns between weekdays and weekends

the dataset with missing values filled in is used for this.The dataset is `activity2`.

add new variable to the `activity2`dataset

```{r}
activity2$date <- as.Date(activity2$date)

activity2$DayType <- ifelse(
    weekdays(activity2$date, abbreviate = FALSE) 
    %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
"Weekdays","Weekend")
         
activity2$DayType <- as.factor(activity2$DayType) 

```

check the levels

```{r}
levels(activity2$DayType)  

```

Review the dataset

```{r}
pander(head(activity2,6))
```

Prepare two new dataset

```{r Weedays and Weekend}
Weekdays <- activity2 %>% filter (DayType == "Weekdays") %>%
    group_by(interval) %>% summarise(Average = mean(steps))

Weekend <- activity2 %>% filter (DayType == "Weekend") %>%
    group_by(interval) %>% summarise(Average = mean(steps))

```

Review the two new dataset

```{r}
pander(head(Weekdays,3))

pander(head(Weekend,3))

```

Create two timeserie 

```{r}
WeekdaysTimeSerie <- ggplot(Weekdays,aes(interval,Average))+
    geom_line()+ylab("Number of Steps")+xlab("5-Minutes Interval")+
    ggtitle("Weekdays")

WeekendTimeSerie <- ggplot(Weekend,aes(interval,Average))+
    geom_line()+ylab("Number of Steps")+xlab("5-Minutes Interval")+
    ggtitle("Weekend")

grid.arrange(WeekdaysTimeSerie,WeekendTimeSerie,nrow = 2)

```





